Carrier phase and amplitude estimation for phase shift keying with unequal error protection

### 2012-12-06 other/uep/paper.tex

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-phase shift keying, and consider specifically the case where a number of different constellations sizes $M$ are used simultaneously.  This scenario is common in practical communications systems employing what is called \emph{unequal error proection}.  We describe an algorithm to compute the least squares estimators of carrier phase and amplitude that requires $O(L \log L)$ arithmetic operations, where $L$ is the number of recieved symbols.   
\end{abstract}
\begin{IEEEkeywords}
Coherent detection, noncoherent detection, phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

In passband communication systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation (amplitude change).  These effects must be compensated for at the receiver. In this paper we assume that the time offset has been previously handled, and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK) and $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M \geq 1$ is the size of the constellation.  We assume that some of the transmitted symbols are \emph{pilot symbols} known to the receiver and the remainder are information carrying \emph{data symbols} with phase that is unknown to the receiver.  So,
\[
s_i = \begin{cases}
p_i & i \in P \\
d_i & i \in D,
\end{cases}
\]
where $P$ is the set of indices describing the position of the pilot symbols $p_i$, and $D$ is a set of indices describing the position of the data symbols $d_i$.  The sets $P$ and $D$ are disjoint, i.e. $P \cap D = \emptyset$, and $L = \abs{P \cup D}$ is the total number of symbols transmitted.  The data symbols are further separated into subsets according to the size of the constellation used to modulate each symbol.  Let $D_2,D_3,\dots$ denote a partition of the indices in $D$ such that $D_2$ is the set of indices corresponding to data symbols modulated with $2$-PSK (i.e BSPK), and $D_3$ is the set of symbols modulated with $3$-PSK, and $D_4$ is the set of symbols modulated with $4$-PSK (i.e. QPSK) and so on.  Let $G$ be the set of integers for which $D_m$ is not empty, i.e. $D_m \neq \emptyset$ whenever $m \in G$.  We have 
\[
D = \bigcup_{m=2}^{\infty}D_m = \bigcup_{m\in G} D_m.
\]
In practical settings the number of different constellations used $\abs{G}$ is small.   A practical example might be that BPSK symbols are used for important symbols and QPSK symbols are used for less important symbols.  In this example $G = \{2,4\}$ and $\abs{G} = 2$.

We assume that time offset estimation has been performed and that $L$ noisy $M$-PSK symbols are observed by the receiver.  The received signal is then,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the reciever and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{equation}\label{eq:SSdefn}
\begin{split}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a s_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2, 
%\\ &= \sum_{i \in P} \abs{ y_i - a s_i }^2 + \sum_{m \in G} \sum_{i \in D_m} \abs{ y_i - a d_i }^2
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is additive white and Gaussian.  However, as we will show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} where no pilot symbols exist ($P = \emptyset$ where $\emptyset$ is the empty set) and where only one constellation size is used, i.e., $G = \{M\}$ contains precisely one element.  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is the so called \emph{non-data aided}, sometimes also called \emph{non-decision directed}, estimator based on the paper of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the recieved signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{A}$ where $\angle$ denotes the complex argument and
\[
A = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\]
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat is that it is not obvious how pilot symbols or data symbols using mulitple constellations should be included in this estimator.  % One approach is to compute
% \[
% A = \frac{1}{L} \sum_{i \in P} F(\abs{y_i}) \tfrac{y_i}{\abs{y_i}} p_i^* +  \frac{1}{L} \sum_{m \in G} \sum_{i \in D_m} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^m
% \]
% and choose the estimate of carrier phase to be $\tfrac{1}{K}\angle{A}$ where
% \[
% K = \begin{cases}
% 1 & \abs{P} > 0 \\
% \min(G) & \text{otherwise},
% \end{cases}
% \]
% is equal to $1$ when pilot symbols exists, and is otherwise equal to the size of the smallest constellation $\min(G)$.  The simulations we present in Section~\ref{sec:simulations} suggest that the accuracy of this estimator is poor when compared to the least squares estimator.  %This problem does not occur with the least square estimator.
This problem does not occur with the least squares estimator.

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Swelden's considered only the noncoherent setting and where only one constellation size is used, but we show in Section~\ref{sec:least-squar-estim}~that Mackenthun's algorithm can be modified to include pilot symbols and multiple constellation sizes. %Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

@@

In this section we derive an algorithm to compute the least squares estimator of the carrier phase and amplitude.  Our algorithm is motivated by the algorithm of Mackenthun~\cite{Mackenthun1994} who considered the noncoherent setting and where only one constellation size is used for the data symbols.  For the purpose of anaylsing computational complexity, we will assume that the number of data symbols $\abs{D}$ is proportional to the total number of symbols $L$, so that, for example, $O(L) = O(\abs{D})$.  In this case Mackentun's algorithm requires $O(L \log L)$ arithmetic operations.

@@

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^*, \label{eq:SS}
\end{align}
where $*$ denotes the complex congugate.  Fixing the data symbols $\{d_i, i \in D\}$, differentiating with respect to $a^*$, and setting the resulting equation to zero we find the least squares estimator of $a_0$ as a function of $\{d_i, i \in D\}$,
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to surpress the argument $(\{ d_i, i \in D\})$ for notational clarity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:SS} we obtain the $SS$ conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ is a constant.  Observe that given candidate values for the data symbols, we can compute the corresponding $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  Let 
\[
H = \sum_{m\in G}m\abs{D_m} = O(L).
\]
It turns out that there are atmost $H$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

@@

Let 
\[
f(\theta) = \{ \hat{d}_i(\theta), i \in D \}
\]
be a function mapping the interval $[0, 2\pi)$ to a sequence of phase shift keyed symbols indexed by the elements of $D$.  Observe that $f(\theta)$ is piecewise continuous.  The subintervals of $[0, 2\pi)$ over which $f(\theta)$ remains contant are determined by the values of $\{z_i, i \in D\}$.  Let
\[
S = \{ f(\theta) \mid \theta \in [0, 2 \pi) \}
\]
be the set of all sequences $f(\theta)$ as $\theta$ varies from $0$ to $2\pi$.  If $\hat{\theta}$ is the least squares estimator the phase then $S$ contains the sequence $\{ \hat{d}_i(\hat{\theta}), i \in D \}$ corresponding to the least squares estimator of the data symbols, i.e., $S$ contains the minimiser of~\eqref{eq:SSdatasymbols}.  Observe from~\eqref{eq:uicombos} that there are atmost $H = O(L)$ sequences in $S$, because there are $m$ distinct values of $\hat{d}_i(\theta)$ for each $i \in D_m$.

The sequences in $S$ can be enumerated as follows.  Let 
\[
T = \bigcup_{m\in G} \bigcup_{ i \in D_m }\bigcup_{k=1}^{m}\left\{ (z_i + \tfrac{\pi(2k - 1)}{m}, i, m) \right\}
\]
be a set of triples with first element a real number, second element from $D$ and third element from $G$.  The number of elements in $T$ is $H = \abs{T} = O(L)$.  Let $t_1,\dots,t_H$ be an enumeration of the triples from $T$ sorted in accending order of the first element in the triple, that is, if $t_{k} = (t_{k1},t_{k2},t_{k3})$ then $t_{i1} \leq t_{k1}$ whenever $i < k$.  Put 
\[
\sigma(k) = t_{k2} \qquad \text{and} \qquad m(k) = t_{k3}
\]
so that $\sigma(1), \dots, \sigma(H)$ and $m(1),\dots,m(H)$ correspond respectively with the second and third elements from the triples $t_1,\dots,t_H$.
%So $a_1,\dots,a_H$ is a sequence of real numbers, $b_1,\dots,b_H$ is a sequence of integers from $D$ and $c_1,\dots,c_H$ is a sequence

@@

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is a constant, independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k}$ be the sequence $\{g_{ik}, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{k}^*,
\]
where $g_{k} e_{k}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/m(k)}$.  Now,
\[
Y_1 = B + \sum_{i \in D} g_{1i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_2 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/m(1)} - 1)g_{1\sigma(1)} + \sum_{i \in D} g_{1i} \\
&= Y_0 + (e^{j2\pi/m(1)} - 1)g_{1\sigma(1)},
\end{align*}
In general,
\[
Y_{k+1} = Y_k + (e^{j2\pi/m(k)} - 1) g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from it predecessor $Y_{k-1}$ in a constant number of operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in a constant number of operations using~\eqref{eq:SSfk}.  Let $\hat{k} = \arg\min SS(f_k)$.  The least squares estimator of the complex amplitude is then computed according to~\eqref{eq:hata},
\begin{equation}\label{eq:ahatYhat}
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\end{equation}
Pseudocode is given in Algorithm~\ref{alg:loglinear}.   The function $\operatorname{sort}$ on line~\ref{alg_sortindices} return the triple $t_1,\dots,t_H$ so that the first element in the triples is in ascending order.   This requires $O(L \log L)$ operations.  The $\operatorname{sort}$ function is the primary bottleneck in this algorithm when $L$ is large.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all require $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

\begin{algorithm}[t] \label{alg:loglinear}
\SetAlCapFnt{\small}
\SetAlTitleFnt{}
\caption{Mackenthun's algorithm with pilot symbols}
\DontPrintSemicolon
\KwIn{$\{y_i, i \in P \cup D \}$}
$c = 1$ \;
\For{$m \in G$ \nllabel{alg_loop_setup}}{
\For{$i \in D_m$ }{
$\phi = \angle{y_i}$ \;
$u = \round{\phi}_m $ \;
$g_i = y_i e^{-j u}$ \;
$z  = \phi - u$ \;
\For{$k = 1, \dots, m$ }{ 
$t_{c} =  (z + \tfrac{\pi(2k - 1)}{m}, i, m)$ \;
$c = c + 1$ \;
} 
}
}
$\operatorname{sort}(t_1,\dots,t_H)$ \nllabel{alg_sortindices} \;
$Y = \sum_{i \in P} y_i p_i^* + \sum_{i \in D} g_i $ \nllabel{alg_Y}\;
$\hat{a} = \frac{1}{L} Y$ \;
$\hat{Q} = \frac{1}{L}\abs{Y}^2$ \nllabel{alg_Q} \;
\For{$k = 1,\dots,H$ \nllabel{alg_loop_search}}{
$Y = Y + (e^{j2\pi/m(k)} - 1) g_{\sigma(k)}$ \;
$g_{\sigma(k)} = e^{j2\pi/m(k)} g_{\sigma(k)} $\;
$Q = \frac{1}{L}\abs{Y}^2$\;
\If{$Q > \hat{Q}$}{
 	$\hat{Q} = Q$ \;
 	$\hat{a} =  \frac{1}{L} Y$ \;
 }
}
\Return{$\hat{a}$ \nllabel{alg_return}}
\end{algorithm}

@@

\section{Conclusion}

We considered least squares estimators of carrier phase and amplitude from noisy communications signals that contain both pilot signals, known to the receiver, and data signals, unknown to the receiver.  We focused on $M$-ary phase shift keying constellations.  The least squares estimator can be computed in $O(L\log L)$ operations using a modification of an algorithm due to Mackenthun~\cite{Mackenthun1994}, and is the maximum likelihood estimator in the case that the noise is additive white and Gaussian.

%We showed, under some conditions on the distribution of the noise, that the phase estimator $\hat{\theta}$ is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator $\hat{\rho}_0$ is biased, and converges to $G(0)\rho_0$.  This bias is large when the signal to noise ratio is small.  It would be interesting to investigate methods for correcting this bias.  A method for estimating $G(0)$ at the receiver appears to be required.

Monte Carlo simulations were used to assess the performance of the least squares estimator.  Interestingly, when the SNR is small, it is counter productive to use the data symbols to estimate the phase (Figure~\ref{fig:plotphaseQPSKmultL}).  This suggests the use of a weighed objective function, which would be an interesting topic for future research.

>>>>

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-ary phase shift keying, and consider specifically the case where a number of different constellations sizes $M$ are used simultaneously.  This scenario is common in practical communications systems employing what is called \emph{unequal error protection}.  We describe an algorithm to compute the least squares estimators of carrier phase and amplitude that requires $O(L \log L)$ arithmetic operations, where $L$ is the number of received symbols.   
\end{abstract}
\begin{IEEEkeywords}
Coherent detection, noncoherent detection, phase shift keying, unequal error protection
\end{IEEEkeywords}

@@

In passband communication systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation (amplitude change).  These effects must be compensated for at the receiver. In this paper we assume that the time offset has been previously handled, and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK) and $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M \geq 2$ is the size of the constellation.  We assume that some of the transmitted symbols are \emph{pilot symbols} known to the receiver and the remainder are information carrying \emph{data symbols} with phase that is unknown to the receiver.  So,
\[
s_i = \begin{cases}
p_i & i \in P \\
d_i & i \in D,
\end{cases}
\]
where $P$ is the set of indices describing the position of the pilot symbols $p_i$, and $D$ is a set of indices describing the position of the data symbols $d_i$.  The sets $P$ and $D$ are disjoint, i.e. $P \cap D = \emptyset$, and $L = \abs{P \cup D}$ is the total number of symbols transmitted.  The data symbols are further separated into subsets according to the size of the constellation used to modulate each symbol.  Let $D_2,D_3,\dots$ denote a partition of the indices in $D$ such that $D_2$ is the set of indices corresponding to data symbols modulated with $2$-PSK (i.e BSPK), and $D_3$ is the set of symbols modulated with $3$-PSK, and $D_4$ is the set of symbols modulated with $4$-PSK (i.e. QPSK) and so on.  Let $G$ be the set of integers for which $D_m$ is not empty, i.e. $\abs{D_m} > 0$ if and only if $m \in G$.  We have 
\[
D = \bigcup_{m=2}^{\infty}D_m = \bigcup_{m\in G} D_m.
\]
In practical settings the number of different constellations used $\abs{G}$ is typically small.   A practical example might be that BPSK is used for important symbols and QPSK is used for less important symbols.  In this example $G = \{2,4\}$ and $\abs{G} = 2$.

We assume that time offset estimation has been performed and that $L$ noisy $M$-PSK symbols are observed by the receiver.  The received signal is then,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the receiver and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\begin{equation}\label{eq:SSdefn}
\begin{split}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a s_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2, 
%\\ &= \sum_{i \in P} \abs{ y_i - a s_i }^2 + \sum_{m \in G} \sum_{i \in D_m} \abs{ y_i - a d_i }^2
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is additive white and Gaussian.  However, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} where no pilot symbols exist ($P = \emptyset$ where $\emptyset$ is the empty set) and where only one constellation size is used, i.e., $G = \{M\}$ contains precisely one element.  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is the so called \emph{non-data aided}, sometimes also called \emph{non-decision directed}, estimator based on the paper of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the received signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{A}$ where $\angle$ denotes the complex argument and
\[
A = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\]
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat is that it is not obvious how pilot symbols or data symbols using multiple constellations should be included in this estimator.  % One approach is to compute
% \[
% A = \frac{1}{L} \sum_{i \in P} F(\abs{y_i}) \tfrac{y_i}{\abs{y_i}} p_i^* +  \frac{1}{L} \sum_{m \in G} \sum_{i \in D_m} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^m
% \]
% and choose the estimate of carrier phase to be $\tfrac{1}{K}\angle{A}$ where
% \[
% K = \begin{cases}
% 1 & \abs{P} > 0 \\
% \min(G) & \text{otherwise},
% \end{cases}
% \]
% is equal to $1$ when pilot symbols exists, and is otherwise equal to the size of the smallest constellation $\min(G)$.  The simulations we present in Section~\ref{sec:simulations} suggest that the accuracy of this estimator is poor when compared to the least squares estimator.  %This problem does not occur with the least square estimator.
This problem does not occur with the least squares estimator.

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Swelden's considered only the noncoherent setting and where only one constellation size is used.  We show in Section~\ref{sec:least-squar-estim}~that Mackenthun's algorithm can be modified to include pilot symbols and multiple constellation sizes. %Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

@@

In this section we derive an algorithm to compute the least squares estimator of the carrier phase and amplitude.  Our algorithm is motivated by the algorithm of Mackenthun~\cite{Mackenthun1994} who considered the noncoherent setting and where only one constellation size is used for the data symbols.  For the purpose of analysing computational complexity, we will assume that the number of data symbols $\abs{D}$ is proportional to the total number of symbols $L$, so that, for example, $O(L) = O(\abs{D})$.  In this case our algorithm requires $O(L \log L)$ arithmetic operations.

@@

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^*, \label{eq:SS}
\end{align}
where $*$ denotes the complex conjugate.  Fixing the data symbols $\{d_i, i \in D\}$, differentiating with respect to $a^*$, and setting the resulting equation to zero we find the least squares estimator of $a_0$ as a function of $\{d_i, i \in D\}$,
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to suppress the argument $(\{ d_i, i \in D\})$ for notational clarity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:SS} we obtain the $SS$ conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
and where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ is a constant.  Observe that given candidate values for the data symbols, we can compute the corresponding $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  Let 
\[
H = \sum_{m\in G}m\abs{D_m} = O(L).
\]
It turns out that there are at most $H$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

@@

Let 
\[
f(\theta) = \{ \hat{d}_i(\theta), i \in D \}
\]
be a function mapping the interval $[0, 2\pi)$ to a sequence of phase shift keyed symbols indexed by the elements of $D$.  Observe that $f(\theta)$ is piecewise continuous.  The subintervals of $[0, 2\pi)$ over which $f(\theta)$ remains constant are determined by the values of $\{z_i, i \in D\}$.  Let
\[
S = \{ f(\theta) \mid \theta \in [0, 2 \pi) \}
\]
be the set of all sequences $f(\theta)$ as $\theta$ varies from $0$ to $2\pi$.  If $\hat{\theta}$ is the least squares estimator the phase then $S$ contains the sequence $\{ \hat{d}_i(\hat{\theta}), i \in D \}$ corresponding to the least squares estimator of the data symbols, i.e., $S$ contains the minimiser of~\eqref{eq:SSdatasymbols}.  Observe from~\eqref{eq:uicombos} that there are atmost $H = O(L)$ sequences in $S$, because there are $m$ distinct values of $\hat{d}_i(\theta)$ for each $i \in D_m$.

The sequences in $S$ can be enumerated as follows.  Let 
\[
T = \bigcup_{m\in G} \bigcup_{ i \in D_m }\bigcup_{k=1}^{m}\left\{ (z_i + \tfrac{\pi(2k - 1)}{m}, i, m) \right\}
\]
be a set of triples with first element a real number, second element from $D$ and third element from $G$.  The number of elements in $T$ is $H = \abs{T} = O(L)$.  Let $t_1,\dots,t_H$ be an enumeration of the triples from $T$ sorted in ascending order of the first element in the triple, that is, if $t_{k} = (t_{k1},t_{k2},t_{k3})$ then $t_{i1} \leq t_{k1}$ whenever $i < k$.  Put 
\[
\sigma(k) = t_{k2} \qquad \text{and} \qquad m(k) = t_{k3}
\]
so that $\sigma(1), \dots, \sigma(H)$ and $m(1),\dots,m(H)$ correspond respectively with the second and third elements from the triples $t_1,\dots,t_H$.
%So $a_1,\dots,a_H$ is a sequence of real numbers, $b_1,\dots,b_H$ is a sequence of integers from $D$ and $c_1,\dots,c_H$ is a sequence

@@

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is a constant, independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k}$ be the sequence $\{g_{ik}, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{k}^*,
\]
where $g_{k} e_{k}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/m(k)}$.  Now,
\[
Y_1 = B + \sum_{i \in D} g_{1i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_2 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/m(1)} - 1)g_{1\sigma(1)} + \sum_{i \in D} g_{1i} \\
&= Y_0 + (e^{j2\pi/m(1)} - 1)g_{1\sigma(1)},
\end{align*}
In general,
\[
Y_{k+1} = Y_k + (e^{j2\pi/m(k)} - 1) g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from it predecessor $Y_{k-1}$ in a constant number of operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in a constant number of operations using~\eqref{eq:SSfk}.  Let $\hat{k} = \arg\min_{k\in\{1,\dots,H\}} SS(f_k)$.  The least squares estimator of the complex amplitude is then computed according to~\eqref{eq:hata},
\begin{equation}\label{eq:ahatYhat}
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\end{equation}
Pseudocode is given in Algorithm~\ref{alg:loglinear}.   The function $\operatorname{sort}$ on line~\ref{alg_sortindices} return the triples $t_1,\dots,t_H$ so that the first element is in ascending order.   This requires $O(H \log H) = O(L \log L)$ operations.  The $\operatorname{sort}$ function is the primary bottleneck in this algorithm when $L$ is large.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all require $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

\begin{algorithm}[t] \label{alg:loglinear}
\SetAlCapFnt{\small}
\SetAlTitleFnt{}
\caption{Mackenthun's algorithm with pilot symbols and multiple constellations sizes}
\DontPrintSemicolon
\KwIn{$\{y_i, i \in P \cup D \}$}
$c = 1$ \;
\For{$m \in G$ \nllabel{alg_loop_setup}}{
\For{$i \in D_m$ }{
$\phi = \angle{y_i}$ \;
$u = \round{\phi}_m $ \;
$g_i = y_i e^{-j u}$ \;
$z  = \phi - u$ \;
\For{$k = 1, \dots, m$ }{ 
$t_{c} =  (z + \tfrac{\pi(2k - 1)}{m}, i, m)$ \;
$c = c + 1$ \;
} 
}
}
$\operatorname{sort}(t_1,\dots,t_H)$ \nllabel{alg_sortindices} \;
$Y = \sum_{i \in P} y_i p_i^* + \sum_{i \in D} g_i $ \nllabel{alg_Y}\;
$\hat{a} = \frac{1}{L} Y$ \;
$\hat{Q} = \frac{1}{L}\abs{Y}^2$ \nllabel{alg_Q} \;
\For{$k = 1,\dots,H$ \nllabel{alg_loop_search}}{
$Y = Y + (e^{j2\pi/m(k)} - 1) g_{\sigma(k)}$ \;
$g_{\sigma(k)} = e^{j2\pi/m(k)} g_{\sigma(k)} $\;
$Q = \frac{1}{L}\abs{Y}^2$\;
\If{$Q > \hat{Q}$}{
 	$\hat{Q} = Q$ \;
 	$\hat{a} =  \frac{1}{L} Y$ \;
 }
}
\Return{$\hat{a}$ \nllabel{alg_return}}
\end{algorithm}

@@
